---
title: "DataPreprocessing2"
author: "Mohammed Alrezq"
date: "2024-10-19"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries 
```{r}
#install.packages("tidyverse")
library(tidyverse)
#install.packages("cli")
library(cli)
#install.packages("rlang")
library(rlang)
library(psych)# for factor analysis 
library(psychTools)
library(ggplot2)
library (descstat)
library(dplyr)
library(REdaS)# to perform Bartlett’s Test of Sphericity before FA 
library(GPArotation) # required by FA 
library(GGally)
library(lavaan)# used for CFA and SEM modeling 
library(lavaanPlot)
library(semPlot)# also use to build graph for CFA, and SEM
library(caret)

```


```{r}
#########Load the data ######################

data <- read.csv("AEMS-2019.csv")
```


```{r}
###################Themes
plotTheme <- function() {
  theme(
    plot.title = element_text(size = 14, family = "sans", face = "plain", hjust = 0),
    plot.subtitle=element_text(size = 11, family = "sans", hjust = 0),
    plot.caption=element_text(size = 10, family = "sans", face = "italic", hjust = 0), 
    axis.title.x = element_text(size = 10, family = "sans", face = "plain", hjust = 1, vjust = -0.5),
    axis.title.y = element_text(size = 10, family = "sans", face = "plain", hjust = 1, vjust = 1),
    axis.text = element_text(size = 9, family = "sans", face = "plain"),
    panel.background = element_blank(),
    panel.grid.minor = element_line(colour = "gray"),
    panel.grid.major = element_line(colour = "gray"),
    axis.ticks = element_blank(),
    legend.title = element_text(size = 10, family = "sans"),
    legend.text = element_text(size = 9, family = "sans"),
    axis.line = element_blank()
  )
}
```



```{r}
# Exploratory Data Analysis 

#describe(data)
# source:
#Variable Name: Source
#1 = School Student
#2 = School Teacher
#3 = University Instructor
#4 = University Student

# to count how many responses by source type 
data%>% count(data$Source)

# to see how many respondent by group (1=school students, 2= teacher, 3=Unv. instructor, 4= Univ. students)
ggplot(data, aes(x=Source))+
  geom_bar(color="darkblue", fill="lightblue")+
  labs(
    title = "Responses Per Participant Type",
    x = "1= School Student, 2 = School Teacher, 3= University Instructor, 4= University Student",
    y = "Count",
    color = "Geschlecht", 
    
  )
table(data$Source)
```


```{r}
library(naniar)# used to replace missing values (e.g. 99,88, etc) with NAs 
# summary package for descriptive stat
#library(summarytools)
```



# High school students: selection and NA investigation 
```{r}
school_student<- data[which(data$Source == "1"),]# select school students only 
names(school_student)
summary(school_student)

#The rest of the dataset (W/O Religion variable)
school_student_rest<- select(school_student,c(1:10, 12:160))

# define NAs values for the rest of the dataset
NA_values<- c("99", "88", "9")
school_studnet_rest_NA<- school_student_rest %>%
  replace_with_na_all(condition = ~.x %in% NA_values)

# Religion variable 
school_studnt_religion<- select(school_student, 11)
#dfSummary(school_studnt_religion)
school_studnt_religion_NA<- school_studnt_religion %>% replace_with_na(replace = list(Religion = 99))
summary(school_studnt_religion_NA)


# Combine the entire dataset with NAs
school_studnet_NA<- cbind(school_studnt_religion_NA, school_studnet_rest_NA)
names(school_studnet_NA)



sapply(school_studnet_NA, function(X) sum(is.na(X)))  # count na per variable

most_missing <- select(school_studnet_NA, c(1:20)) # visualizing the NA for the variable that has missing more than 10% 
gg_miss_var(most_missing, show_pct = TRUE)
```


#Remove  variables for teachers and univ. students and only keep school studnets variables 
```{r}
names(school_studnet_NA)
school_student_Variabler<- school_studnet_NA[c(-2, -5:-6, -10:-11,-14:-16, -97:-123)]# deleted:source names, resp. ID, school type, degree year, education, relationship status, children, work experience, and other construct not related to student 

names(school_student_Variabler)

most_missing_schoolStudent <- select(school_student_Variabler, c(1:20)) # visualizing the NA for the variable that has missing more than 10% 
gg_miss_var(most_missing_schoolStudent, show_pct = TRUE)

# Remove Grade and use age as an alternative. also, remove all other NAs in countery, ethincity, etc 

school_student_cleaning1 <- school_student_Variabler[c(-6)]
names(school_student_cleaning1)

# Visulize after removing Grade 
most_missing_clean1 <- select(school_student_cleaning1, c(1:35)) # visualizing the NA for the variable that has missing more than 10% 
gg_miss_var(most_missing_clean1, show_pct = TRUE)

```


#Omit NAs from school student data
```{r}
school_student_cleaning2 <- na.omit(school_student_cleaning1)
#summary(school_student_cleaning2)

# to check which row and how many obs.  =8 for the two variables with 8
which(school_student_cleaning2$SR_Problems==8)
which(school_student_cleaning2$SB_BeMyself==8)

# remove "na = 8" from SB.bemyself, and SR.problems

schoolData<- filter(school_student_cleaning2, 
                  !SR_Problems==8, 
                  !SB_BeMyself==8)

```


```{r}
#write.csv(schoolData, "school_data.csv")
```


# revers coding for some CIO Items
```{r}

reverse_cols <- c ("CIO_SelfDepend", 
                   "CIO_SelfDependMost",
                   "CIO_OwnThing",
                   "CIO_Identity",
                   "CIO_JobBetter",
                   "CIO_Competition",
                   "CIO_BetterTense",
                   
"Gratitude_NotMuch",
"Gratitude_Time",
"Empathy_Waste",
"Empathy_Difficult",
"SR_GoalSettingHard",
"SR_GoalPlanTrouble",
"SR_Distracted",
"SR_Trouble",
"SR_Decisions",
"SR_Change",
"SR_Problems",
"SR_Focus",
"SR_Mistakes",
"SB_DifferentSchool",
"SB_NotInterested",
"SB_FeelDifferent",
"SB_DontBelong",
"SB_AcceptanceHard",
"MM_NoPurpose",
"SE_Insecure",
"SE_DontHandle",
"SE_NotCapable",
"SE_RarelyAchieve",
"SE_NewGiveUp",
"SE_AvoidDifficult",
"SE_NotTryComplicated")

schoolData[,reverse_cols] <- 5 - schoolData[,reverse_cols]
#str(schoolData)
#summary(schoolData)
```


10/19/2024
```{r}
################################filter 1#########################333
names(school)


school_less24<- subset.data.frame(schoolData, schoolData$Age<= 2)
table(school_less24$Country)
names(school_less24)

# Rename Country to Region
school_rename<- rename(school_less24,
       Region= "Country")
names(school_rename)

table(school_rename$Region)


###############################filter 2##########################333

# Further filtiring the data based on age (less than 18 yrs only)

names(school_model)
str(school_model)
head(school_model)
table(school_model$Age)
# filter datset based on age less than 18 yrs
#school_model<- school_model[-c(14,7)]# remove age and religious/spirtuality
school_rename<- school_rename[which(school_rename$Age == "1"),]

# remove age column 
school_rename<- school_rename[-c(14)]# remove age 
 

school_rename$Region<- as.factor(school_rename$Region)
school_rename$Gender<- as.factor(school_rename$Gender)

names(school_rename)
```




#Grouping itmes to single construct based on the mean  of each row 

```{r}
#package for correlation 
#install.packages("GGally")
library(GGally)
names(schoolData)
library(corrplot)

#SB
SB_data<- select(schoolData,c(65:82))
str(SB_data)

ggcorr(SB_data, method = c("everything", "pearson"))
corPlot(SB_data, cex = 0.5)

SB_mean<- SB_data %>% as_tibble() %>% 
   mutate(mean_SB = rowMeans(across(where(is.numeric))))
SB_mean

#PS
ps_data<- select(schoolData,c(43:54))
str(ps_data)
corPlot(ps_data, cex = 0.5)
ps_mean<- ps_data %>% as_tibble() %>% 
   mutate(mean_ps = rowMeans(across(where(is.numeric))))
ps_mean

#ER
ER_data<- select(schoolData,c(94:101))
corPlot(ER_data, cex = 0.5)

ER_mean<- ER_data %>% as_tibble() %>% 
   mutate(mean_ER = rowMeans(across(where(is.numeric))))
ER_mean

#Empathy 
Empathy_data <-  select(schoolData,c(102:108))
corPlot(Empathy_data, cex = 0.5)

Empathy_mean<- Empathy_data %>% as_tibble() %>% 
   mutate(mean_Empathy = rowMeans(across(where(is.numeric))))
Empathy_mean

#RS
RS_data<- select(schoolData,c(83:87))
corPlot(RS_data, cex = 0.5)

RS_mean<- RS_data %>% as_tibble() %>% 
   mutate(mean_RS = rowMeans(across(where(is.numeric))))
RS_mean

#SR
SR_data<- select(schoolData,c(109:124))
corPlot(SR_data, cex = 0.5)

SR_mean<- SR_data %>% as_tibble() %>% 
   mutate(mean_SR = rowMeans(across(where(is.numeric))))
SR_mean

#Forg
forgive_data<- select(schoolData,c(8:16))
corPlot(forgive_data, cex = 0.5)

forgive_mean<- forgive_data %>% as_tibble() %>% 
   mutate(mean_forgive = rowMeans(across(where(is.numeric))))
forgive_mean

#MM
MM_data<- select(schoolData,c(55:64))
corPlot(MM_data, cex = 0.5)

MM_mean<- MM_data %>% as_tibble() %>% 
   mutate(mean_MM = rowMeans(across(where(is.numeric))))
MM_mean
#SE
SE_data<- select(schoolData,c(31:42))
corPlot(SE_data, cex = 0.5)

SE_mean<- SE_data %>% as_tibble() %>% 
   mutate(mean_SE = rowMeans(across(where(is.numeric))))
SE_mean

#Gratidute 
gratitude_data<- select(schoolData,c(88:93))
corPlot(gratitude_data, cex = 0.5)

gratitude_mean<- gratitude_data %>% as_tibble() %>% 
   mutate(mean_gratitude = rowMeans(across(where(is.numeric))))
gratitude_mean

#CIO
CIO_data<- select(schoolData,c(17:30))
corPlot(CIO_data, cex = 0.5)

CIO_mean<- CIO_data %>% as_tibble() %>% 
   mutate(mean_CIO = rowMeans(across(where(is.numeric))))


```


#Combined the mean into one dataset 
```{r}

mean_school.Data<- cbind(CIO_mean$mean_CIO, gratitude_mean$mean_gratitude,MM_mean$mean_MM, SE_mean$mean_SE, forgive_mean$mean_forgive, SR_mean$mean_SR, RS_mean$mean_RS,ER_mean$mean_ER,Empathy_mean$mean_Empathy, ps_mean$mean_ps, SB_mean$mean_SB)

mean_school.Data <- as.data.frame(mean_school.Data)

```

#Rename variables/construct
```{r}
names(mean_school.Data)

mean_school.Data<- rename(mean_school.Data,
       CIO= "V1",
       GRAT= "V2",
       MEANMK= "V3",
       SEFF= "V4",
       FORG= "V5",
       SREG= "V6",
       RELSPIR = "V7",
       EMOREG= "V8", 
       EMP= "V9",
       PSOLV= "V10",
       SOB="V11") 
names(mean_school.Data)


```
 
# Final data for modeling 
```{r}

#add deographic data 
names(schoolData)
Demo_schoo.data<- select(schoolData, c(3:5))
names(Demo_schoo.data)
 
school <- cbind (mean_school.Data, Demo_schoo.data)# the entire data set (including demographic and construct)
str(school)
```


#Remove participants older than 24 years 
```{r}
names(school)


school_less24<- subset.data.frame(school, school$Age <= 2)
table(school_less24$Country)
names(school_less24)

# Rename Country to Region
school_rename<- rename(school_less24,
       Region= "Country")
names(school_rename)

table(school_rename$Region)

```


# Recoding region for demographic data in the entire data set
```{r}
library(tidyverse)
#the below code did not run on 06/14/2024 and i used the code under it instead but i need to further check why recode fucntion did not work
# school_recoding <- school_rename %>%
#                    mutate(Region = recode(Region, 
#                                           "15" = 1,
#                                           "1" = 1,
#                                           "14" = 2,
#                                           "2"= 2,
#                                           "4"= 3,
#                                           "7" = 3, 
#                                           "5" = 4,
#                                           "13" = 4,
#                                           "6" = 5,
#                                           "11"= 5,
#                                           "12" = 5, 
#                                           "8" = 0,
#                                           "10"= 0))
# 
# 
# 
# names(school_recoding)
# table(school_recoding$Region)


# The belwo code for recoding Region values was repalced by above code in 06/14/2024
school_recoding <- school_rename %>%
  mutate(Region = case_when(
    Region %in% c("15", "1") ~ 1,
    Region %in% c("14", "2") ~ 2,
    Region %in% c("4", "7") ~ 3,
    Region %in% c("5", "13") ~ 4,
    Region %in% c("6", "11", "12") ~ 5,
    Region %in% c("8", "10") ~ 0,
    TRUE ~ as.numeric(Region)  # Default case if none of the above match
  ))






```


# Remove demographic varibales as suggested by stakeholder 
```{r}


school_model<- (school_recoding)
names(school_model)
str(school_model)
```
#Variable visulization after taking mean  
```{r}
str(school_model)
par(mfrow=c(2,2))
hist(school_model$CIO)
hist(school_model$GRAT)
hist(school_model$MEANMK)
hist(school_model$SEFF)
hist(school_model$FORG)
hist(school_model$SREG)
hist(school_model$RELSPIR)
hist(school_model$EMOREG)
hist(school_model$EMP)
hist(school_model$PSOLV)
hist(school_model$SOB)


# Summary statistics for all variables table 
#Source:https://cran.r-project.org/web/packages/vtable/vignettes/sumtable.html
library(vtable)
#st(school_model)

sumtable(school_model,vars = c('CIO','Grat','MM','Self.Efc','Self_Reg', 'Relg.Spit','ER','EMP','PR.S', 'Self.Bel', 'Forg'),
         summ=c('notNA(x)',
                'mean(x)',
                'median(x)',
                'sd(x)',
                'min(x)',
                'max(x)'))
table(school_model$Region)


```




# Use Cronbach’s Alpha to check the the items for each factor (10/19/2024)
```{r}
library(psych)
names(school_rename)

forgive <- alpha(school_rename[,c(8:15)])  

CIO <- alpha(school_rename[,c(16:27, 29)]
             #, check.keys = TRUE# within the code, we can run this check for code reverse, any variabels with negative sign then it is reversed
             )  


SE <- alpha(school_rename[,c(30:41)])  


PS <- alpha(school_rename[,c(42:53)])  

MM <- alpha(school_rename[,c(54:63)])  

SB <- alpha(school_rename[,c(64:81)])  

SB <- alpha(school_rename[,c(64:81)])  

RS <- alpha(school_rename[,c(82:86)])  

Gratitude <- alpha(school_rename[,c(87:92)])  

ER <- alpha(school_rename[,c(93:100)])  

Empth <- alpha(school_rename[,c(101:107)])  
                                  
SR <- alpha(school_rename[,c(108:123)])  



      

```


#combining items into a construct based on mean
```{r}
# Combine multiple items to create the construct using column indices
data <- data %>%
  mutate(Construct_A = rowMeans(select(., 1:3), na.rm = TRUE))
```



```{r}
library(ggplot2)# for correlqtion plot 
library(ggExtra)
library(corrplot)
# Corrplot source: https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html
```


 

# this is aqfter author feedback- 10/19/2024
```{r}
boxplot(school_model)
names(school_model)
school_model_revision <- school_model |> 
  select(-Region)

# Load the required libraries
library(ggplot2)
library(dplyr)
library(tidyr)


# Convert data to long format
data_long <- school_model_revision %>%
  pivot_longer(cols = -Gender, names_to = "Variable", values_to = "Value")

# Create the boxplot
ggplot(data_long, aes(x = Value, y = Variable, fill = Gender)) +
  geom_boxplot() +
  facet_wrap(~ Gender, ncol = 2) +
  labs(
       x = "Scale",
       y = "Variables") +
  scale_fill_manual(values = c("0" = "Female", "1" = "Male"), 
                    labels = c("0" = "Female", "1" = "Male")) +
  theme_minimal()

```



```{r, include=FALSE}
library(mlbench)
library(caret)# an aggregator package for performing many machine learning models
library(e1071)
library(lime)

library(rsample)      # data splitting 

library(ranger)       # a faster implementation of randomForest
library(h2o)          # an extremely fast java-based platform
library(rsq)
```

# Further filtiring the data based on age (less than 18 yrs only)
```{r}
names(school_model)
str(school_model)
head(school_model)
table(school_model$Age)
# filter datset based on age less than 18 yrs
#school_model<- school_model[-c(14,7)]# remove age and religious/spirtuality
school_model<- school_model[which(school_model$Age == "1"),]

# remove age column 
school_model<- school_model[-c(14)]# remove age 
 

school_model$Region<- as.factor(school_model$Region)
school_model$Gender<- as.factor(school_model$Gender)
```


# Split data for regression 
```{r}
school_modelReg<- school_model# to make the name of the data sepcific for regression 
str(school_modelReg)

set.seed(12346)
ind <- sample(2, nrow(school_modelReg), replace = T, prob = c(0.8, 0.2))
train_reg <- school_modelReg[ind == 1,]
test_reg <- school_modelReg[ind == 2,]


```

#Package used for regression 
```{r, include=FALSE}
library(randomForest)

library(DAAG)
library(MASS)
library(compositions)

library(car)
library(PerformanceAnalytics)
library(olsrr)# to run all possible soltuions for regression: forward, backword, both
library(performance)# to cehkc linear regression assumption 
library(glmnet)# for lasso regression 
install.packages("glmnet", repos = "http://cran.us.r-project.org")
```


